PANCHANGE_data_preprocessing

### 2020-08-14 PANCHANGE_data_preprocessing2020-07-15.Rmd
Finished adapting respiratory section. Confirmed the script runs fully for NBR cohort.

# Specify cohort
```{r Specify cohort}
GLAD = FALSE
EDGI = TRUE
NBR = FALSE
RAMP = FALSE
```

@@

  edgi.mhd2.raw.id <- edgi.mhd2.raw %>%
    drop_na(externalDataReference) %>% # Drop NAs
    distinct(externalDataReference, .keep_all = TRUE) %>% # Remove duplicates based on ID
    separate(externalDataReference, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
    mutate(ID = as.numeric(ID)) %>%
    select(
      Sample, # Sample
      ID, # ID
    mhd.depression = demographics.depression,
    mhd.pregnancy_depression = demographics.pregnancy_depression,
    mhd.premenstrual_dysphoric_disorder_pmdd = demographics.premenstrual_dysphoric_disorder_pmdd,
    mhd.anxiety_nerves_or_generalised_anxiety_disorder = demographics.anxiety_nerves_or_generalised_anxiety_disorder,
    mhd.social_anxiety_or_social_phobia = demographics.social_anxiety_or_social_phobia,
    mhd.specific_phobia_e.g._phobia_of_flying = demographics.specific_phobia_e.g._fear_of_flying, # fear instead of phobia
    mhd.agoraphobia = demographics.agoraphobia,
    mhd.panic_disorder = demographics.panic_disorder,
    mhd.panic_attacks = demographics.panic_attacks,
    mhd.obsessivecompulsive_disorder_ocd = demographics.obsessivecompulsive_disorder_ocd,
    mhd.obsessive_compulsive_related_disorders = demographics.obsessive_compulsive_related_disorders,
    mhd.mania_hypomania_bipolar_or_manicdepression = demographics.mania_hypomania_bipolar_or_manicdepression,
    mhd.schizophrenia = demographics.schizophrenia,
    mhd.type_psychosis_psychotic_illness = demographics.psychosis_type_psychotic_illness, # word type psychosis changed order compared to GLAD 
#    mhd.personality_disorder_diagnosed = demographics.personality_disorder_diagnosed, # does not exist in EDGI because it is multiple choice in EDGI
    mhd.posttraumatic_stress_disorder_ptsd = demographics.posttraumatic_stress_disorder_ptsd,
    mhd.autism_aspergers_or_autistic_spectrum_disorder = demographics.autism_spectrum_disorder_asd, # different name in EDGI
    ) %>%
    add_column(
      mhd.schizoaffective_disorder = NA_character_
    ) %>%
    add_numeric(., exclude = exclude_cols_mhd2) %>%
    mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
    mutate_if(is.numeric, ~na_if(., -99)) %>%
    mutate_if(is.numeric, ~na_if(., -77)) %>%
    mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
    mutate_if(is.factor, ~na_if(., "Don't know")) %>%
    mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
    mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
    mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
    mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
    mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
    mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

@@

@@

@@

#Distribution of likert scale COVID-19 symptoms in COPING
```{r distribution of COVID symptoms}
#persistent cough
coping.glad.respiratory.raw.id %>%
  ggplot(
    mapping = aes(
      y = respiratory.persistent_cough_numeric 
    )
  ) +
  geom_bar() +
  labs(
    title = "Persistant cough"
  )

#severe fatigue
coping.glad.respiratory.raw.id %>%
  ggplot(
    mapping = aes(
        y = respiratory.feeling_tiredfatigued_numeric 
    )
  ) +
  geom_bar() +
  labs(
    title = "Severe fatigue"
  )

#skipped meals
coping.glad.respiratory.raw.id %>%
  ggplot(
    mapping = aes(
        y = respiratory.skipping_meals_because_you_felt_unwell_numeric
    )
  ) +
  geom_bar() +
  labs(
    title = "Skipped meals"
  )

@@

Results:
Persistent cough: "Mild symptoms" as cut off
Fatigue: "Severe symptoms" as cut off
Skipped meals: "Moderate symptoms" as cut off
```{r proportion of people who report COVID symptoms in Menni et al}
#persistent cough
percentage_pcough <- ((0.5673*6452) + (0.4556*9186) + (0.3017*2434931))/(6452+9186+2434931)
percentage_pcough    # 0.3029762% of Menni et al(2020) have fatigue
quantile(x = coping.glad.respiratory.raw.id$respiratory.persistent_cough_numeric, 
         na.rm = TRUE, 
         probs = 1-percentage_pcough) # this gives 0.3029762% of the sample reporting fatigue when using the "mild symptoms" persistent cough response

#fatigue
percentage_fatigue <- ((0.298*6452) + (0.1556*9186) + (0.0871*2434931))/(6452+9186+2434931) 
percentage_fatigue   # 0.08791204% of Menni et al(2020) have fatigue
quantile(x = coping.glad.respiratory.raw.id$respiratory.feeling_tiredfatigued_numeric, 
         na.rm = TRUE, 
         probs = 1-percentage_fatigue) #this gives 8% of the sample reporting fatigue when using the "severe symptoms" fatigue response

#skipped meals 
percentage_skipped_meals <- ((0.4203*6452) + (0.2493*9186) + (0.1912*2434931))/(6452+9186+2434931) 
percentage_skipped_meals    # 0.192021% of Menni et al(2020) have fatigue
quantile(x = coping.glad.respiratory.raw.id$respiratory.skipping_meals_because_you_felt_unwell_numeric, 
         na.rm = TRUE, 
         probs = 1-percentage_skipped_meals) # this gives 0.192021% of the sample reporting fatigue when using the "moderate symptoms" persistent cough response
```

Creating binary groups for COVID symptoms
```{r creating binary groups for COVID symptoms}
#loss or smell or taste
coping.glad.respiratory.raw.id <- coping.glad.respiratory.raw.id %>%
  mutate(
    respiratory.anosmia_binary =
      if_else(
        respiratory.was_your_sense_of_smell_impaired_numeric == 1 | respiratory.was_your_sense_of_taste_impaired_numeric == 1, 
        1,
        0
      )
  )

#persistent cough
coping.glad.respiratory.raw.id <- coping.glad.respiratory.raw.id %>%
  mutate(
    respiratory.pcough_binary =
      if_else(
        respiratory.persistent_cough_numeric >= 1, #those who score 1 or above (mild or above)
        1,
        0
      )
  )

#fatigue
coping.glad.respiratory.raw.id <- coping.glad.respiratory.raw.id %>%
  mutate(
    respiratory.fatigue_binary =
      if_else(
        respiratory.feeling_tiredfatigued_numeric >= 3, #those who score 3 (severe symptoms)
        1,
        0
      )
  )

#skipped meals
coping.glad.respiratory.raw.id <- coping.glad.respiratory.raw.id %>%
  mutate(
    respiratory.skipped_meals_binary =
      if_else(
        respiratory.skipping_meals_because_you_felt_unwell_numeric >= 2,
        1,
        0
      )
  )
```

Frequencies of binary COVID symptoms
```{r frequencies of COVID binary symptoms}
freq(coping.glad.respiratory.raw.id$respiratory.anosmia_binary)
freq(coping.glad.respiratory.raw.id$respiratory.pcough_binary)
freq(coping.glad.respiratory.raw.id$respiratory.fatigue_binary)
freq(coping.glad.respiratory.raw.id$respiratory.skipped_meals_binary)
```

Rename respiratory dataset joined to main dataset
```{r clean respiratory dataset}
coping.glad.respiratory <- coping.glad.respiratory.raw.id

dat.coping <- left_join(
  x = dat,
  y = coping.glad.respiratory,
  by = "ID"
)
```

@@

```{r how many people expereinced all symptoms}
dat.coping <- dat.coping %>%
  mutate(
    covid_symptoms =
      if_else(
          !is.na(respiratory.anosmia_binary) &
          !is.na(respiratory.pcough_binary) &
          !is.na(respiratory.fatigue_binary) &
          !is.na(respiratory.skipped_meals_binary),
        1,
        0
       )
  )

@@

Probable COVID case including the routing question
```{r}
dat.coping <- dat.coping %>%
  mutate(
    covid_probable_case_incl_screening_numeric =
      case_when(
        respiratory.breath_diarrhoea_fatigue_fever_numeric == 0 ~ 0,
        covid_odds_ratio_50_numeric == 0 ~ 0,
        covid_odds_ratio_50_numeric == 1 ~ 1
      )
  )

>>>>

# Specify cohort
```{r Specify cohort}
GLAD = FALSE
EDGI = FALSE
NBR = TRUE
RAMP = FALSE
```

@@

  edgi.mhd2.raw.id <- edgi.mhd2.raw %>%
    drop_na(externalDataReference) %>% # Drop NAs
    distinct(externalDataReference, .keep_all = TRUE) %>% # Remove duplicates based on ID
    separate(externalDataReference, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
    mutate(ID = as.numeric(ID)) %>%
    select(
      Sample, # Sample
      ID, # ID
    mhd.depression = demographics.depression,
    mhd.pregnancy_depression = demographics.pregnancy_depression,
    mhd.premenstrual_dysphoric_disorder_pmdd = demographics.premenstrual_dysphoric_disorder_pmdd,
    mhd.anxiety_nerves_or_generalised_anxiety_disorder = demographics.anxiety_nerves_or_generalised_anxiety_disorder,
    mhd.social_anxiety_or_social_phobia = demographics.social_anxiety_or_social_phobia,
    mhd.specific_phobia_e.g._phobia_of_flying = demographics.specific_phobia_e.g._fear_of_flying, # fear instead of phobia
    mhd.agoraphobia = demographics.agoraphobia,
    mhd.panic_disorder = demographics.panic_disorder,
    mhd.panic_attacks = demographics.panic_attacks,
    mhd.obsessivecompulsive_disorder_ocd = demographics.obsessivecompulsive_disorder_ocd,
    mhd.obsessive_compulsive_related_disorders = demographics.obsessive_compulsive_related_disorders,
    mhd.mania_hypomania_bipolar_or_manicdepression = demographics.mania_hypomania_bipolar_or_manicdepression,
    mhd.schizophrenia = demographics.schizophrenia,
    mhd.type_psychosis_psychotic_illness = demographics.psychosis_type_psychotic_illness, # word type psychosis changed order compared to GLAD 
#    mhd.personality_disorder_diagnosed = demographics.personality_disorder, # does not exist in EDGI because it is multiple choice in EDGI
    mhd.posttraumatic_stress_disorder_ptsd = demographics.posttraumatic_stress_disorder_ptsd,
    mhd.autism_aspergers_or_autistic_spectrum_disorder = demographics.autism_spectrum_disorder_asd, # different name in EDGI
    mhd.body_dysmorphic_disorder_bdd = demographics.body_dysmorphic_disorder,
    mhd.personality_disorder = demographics.personality_disorder, # routing variable in EDGI is named differently; renamed here to match GLAD
    ) %>%
    add_column(
      mhd.schizoaffective_disorder = NA_character_
    ) %>%
    add_numeric(., exclude = exclude_cols_mhd2) %>%
    mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
    mutate_if(is.numeric, ~na_if(., -99)) %>%
    mutate_if(is.numeric, ~na_if(., -77)) %>%
    mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
    mutate_if(is.factor, ~na_if(., "Don't know")) %>%
    mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
    mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
    mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
    mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
    mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
    mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

@@

++MD: One NBR ID is being converted to NA by the as.numeric(ID) line of code.

@@

```{r coping glad respiratory copy}
if(GLAD == TRUE){
respiratory.raw.id <- coping.glad.respiratory.raw.id
}
```

@@

```{r coping edgi respiratory copy}
if(EDGI == TRUE){
respiratory.raw.id <- coping.edgi.respiratory.raw.id
}
```

## NBR respiratory
```{r nbr coping respiratory cols}
if(NBR == TRUE){
coping.nbr.respiratory.raw <- readRDS(file = paste0(data.raw_path, "/nbr/respiratory_coping_nbr.rds"))
dim(coping.nbr.respiratory.raw)
colnames(coping.nbr.respiratory.raw)
}
```

Select respiratory variables
```{r coping nbr respiratory cleaning}
if(NBR == TRUE){
coping.nbr.respiratory.raw.id <- coping.nbr.respiratory.raw %>%
  drop_na(subjectid) %>% # Drop NAs
  distinct(subjectid, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(subjectid, into = c("Sample", "ID.intm"), sep = 6) %>% # Split ID in Sample and Number
  separate(ID.intm, into = "ID", sep = 7) %>%
  mutate(ID = as.numeric(ID)) %>%
  select(
    Sample, # Sample
    ID, # ID
    respiratory.breath_diarrhoea_fatigue_fever = respiratory.unusual_shortness_breath_diarrhoea,
    respiratory.worst_ill_feel,
    respiratory.persistent_cough,
    respiratory.skipping_meals_because_you_felt_unwell,
    respiratory.was_your_sense_of_smell_impaired,
    respiratory.was_your_sense_of_taste_impaired,
    respiratory.feeling_tiredfatigued,
    respiratory.throat_swab_coronavirus_nose,
    respiratory.throat_swab_test_nose,
    respiratory.coronavirus_antibody_test,
    respiratory.results_antibody_test
  ) %>%
  add_numeric(., exclude = exclude_cols) %>%
  mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
  mutate_if(is.numeric, ~na_if(., -99)) %>%
  mutate_if(is.numeric, ~na_if(., -77)) %>%
  mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
  mutate_if(is.factor, ~na_if(., "Don't know")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

# Inspect dimensions
dim(coping.nbr.respiratory.raw.id)
# Inspect colnames
colnames(coping.nbr.respiratory.raw.id)
#Differences
dim(coping.nbr.respiratory.raw)[1]-dim(coping.nbr.respiratory.raw.id)[1]
}
```

```{r coping nbr respiratory copy}
if(NBR == TRUE){
respiratory.raw.id <- coping.nbr.respiratory.raw.id
}
```

## RAMP respiratory
```{r ramp coping respiratory cols}
if(RAMP == TRUE){
ramp.respiratory.raw <- readRDS(file = paste0(data.raw_path, "/ramp/respiratory_coping_ramp.rds"))
dim(ramp.respiratory.raw)
colnames(ramp.respiratory.raw)
}
```

Select respiratory variables
```{r coping ramp respiratory cleaning}
if(RAMP == TRUE){
ramp.respiratory.raw.id <- ramp.respiratory.raw %>%
  drop_na(externalDataReference) %>% # Drop NAs
  distinct(externalDataReference, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(externalDataReference, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
  mutate(ID = as.numeric(ID)) %>%
  select(
    Sample, # Sample
    ID, # ID
    respiratory.breath_diarrhoea_fatigue_fever,
    respiratory.worst_ill_feel,
    respiratory.persistent_cough,
    respiratory.skipping_meals_because_you_felt_unwell,
    respiratory.was_your_sense_of_smell_impaired,
    respiratory.was_your_sense_of_taste_impaired,
    respiratory.feeling_tiredfatigued,
    respiratory.throat_swab_coronavirus_nose,
    respiratory.throat_swab_test_nose,
    respiratory.coronavirus_antibody_test,
    respiratory.results_antibody_test
  ) %>%
  add_numeric(., exclude = exclude_cols) %>%
  mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
  mutate_if(is.numeric, ~na_if(., -99)) %>%
  mutate_if(is.numeric, ~na_if(., -77)) %>%
  mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
  mutate_if(is.factor, ~na_if(., "Don't know")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

# Inspect dimensions
dim(ramp.respiratory.raw.id)
# Inspect colnames
colnames(ramp.respiratory.raw.id)
#Differences
dim(ramp.respiratory.raw)[1]-dim(ramp.respiratory.raw.id)[1]
}
```

```{r ramp respiratory copy}
if(RAMP == TRUE){
respiratory.raw.id <- ramp.respiratory.raw.id
}
```

#Distribution of likert scale COVID-19 symptoms in COPING
```{r distribution of COVID symptoms}
#persistent cough
respiratory.raw.id %>%
  ggplot(
    mapping = aes(
      y = respiratory.persistent_cough_numeric 
    )
  ) +
  geom_bar() +
  labs(
    title = "Persistant cough"
  )

#severe fatigue
respiratory.raw.id %>%
  ggplot(
    mapping = aes(
        y = respiratory.feeling_tiredfatigued_numeric 
    )
  ) +
  geom_bar() +
  labs(
    title = "Severe fatigue"
  )

#skipped meals
respiratory.raw.id %>%
  ggplot(
    mapping = aes(
        y = respiratory.skipping_meals_because_you_felt_unwell_numeric
    )
  ) +
  geom_bar() +
  labs(
    title = "Skipped meals"
  )

@@

Results:
Persistent cough: "Mild symptoms" as cut off
Fatigue: "Severe symptoms" as cut off
Skipped meals: "Moderate symptoms" as cut off
```{r proportion of people who report COVID symptoms in Menni et al}
#persistent cough
percentage_pcough <- ((0.5673*6452) + (0.4556*9186) + (0.3017*2434931))/(6452+9186+2434931)
percentage_pcough    # 0.3029762% of Menni et al(2020) have fatigue
quantile(x = respiratory.raw.id$respiratory.persistent_cough_numeric, 
         na.rm = TRUE, 
         probs = 1-percentage_pcough) # this gives 0.3029762% of the sample reporting fatigue when using the "mild symptoms" persistent cough response

#fatigue
percentage_fatigue <- ((0.298*6452) + (0.1556*9186) + (0.0871*2434931))/(6452+9186+2434931) 
percentage_fatigue   # 0.08791204% of Menni et al(2020) have fatigue
quantile(x = respiratory.raw.id$respiratory.feeling_tiredfatigued_numeric, 
         na.rm = TRUE, 
         probs = 1-percentage_fatigue) #this gives 8% of the sample reporting fatigue when using the "severe symptoms" fatigue response

#skipped meals 
percentage_skipped_meals <- ((0.4203*6452) + (0.2493*9186) + (0.1912*2434931))/(6452+9186+2434931) 
percentage_skipped_meals    # 0.192021% of Menni et al(2020) have fatigue
quantile(x = respiratory.raw.id$respiratory.skipping_meals_because_you_felt_unwell_numeric, 
         na.rm = TRUE, 
         probs = 1-percentage_skipped_meals) # this gives 0.192021% of the sample reporting fatigue when using the "moderate symptoms" persistent cough response
```

Creating binary groups for COVID symptoms
```{r creating binary groups for COVID symptoms}
#loss or smell or taste
respiratory.raw.id <- respiratory.raw.id %>%
  mutate(
    respiratory.anosmia_binary =
      if_else(
        respiratory.was_your_sense_of_smell_impaired_numeric == 1 | respiratory.was_your_sense_of_taste_impaired_numeric == 1, 
        1,
        0
      )
  )

#persistent cough
respiratory.raw.id <- respiratory.raw.id %>%
  mutate(
    respiratory.pcough_binary =
      if_else(
        respiratory.persistent_cough_numeric >= 1, #those who score 1 or above (mild or above)
        1,
        0
      )
  )

#fatigue
respiratory.raw.id <- respiratory.raw.id %>%
  mutate(
    respiratory.fatigue_binary =
      if_else(
        respiratory.feeling_tiredfatigued_numeric >= 3, #those who score 3 (severe symptoms)
        1,
        0
      )
  )

#skipped meals
respiratory.raw.id <- respiratory.raw.id %>%
  mutate(
    respiratory.skipped_meals_binary =
      if_else(
        respiratory.skipping_meals_because_you_felt_unwell_numeric >= 2,
        1,
        0
      )
  )
```

Frequencies of binary COVID symptoms
```{r frequencies of COVID binary symptoms}
freq(respiratory.raw.id$respiratory.anosmia_binary)
freq(respiratory.raw.id$respiratory.pcough_binary)
freq(respiratory.raw.id$respiratory.fatigue_binary)
freq(respiratory.raw.id$respiratory.skipped_meals_binary)
```

Rename respiratory dataset joined to main dataset

++MD: Do we want to name the RAMP dataset something different, if the algorithm will need to be changed because they have categorical age and gender not sex? (otherwise can remove the if statement and rename the dataset 'dat.full' instead of dat.coping)
```{r clean respiratory dataset}
if(GLAD == TRUE | EDGI == TRUE | NBR == TRUE){
respiratory <- respiratory.raw.id

dat.coping <- left_join(
  x = dat,
  y = respiratory,
  by = "ID"
)
}
```

@@

```{r how many people experinced all symptoms}
dat.coping <- dat.coping %>%
  mutate(
    covid_symptoms =
      if_else(
          !is.na(respiratory.anosmia_binary) &
          !is.na(respiratory.pcough_binary) &
          !is.na(respiratory.fatigue_binary) &
          !is.na(respiratory.skipped_meals_binary),
        1,
        0
       )
  )

@@

Probable COVID case including the routing question
```{r recoding probable covid to account for screening question}
dat.coping <- dat.coping %>%
  mutate(
    covid_probable_case_incl_screening_numeric =
      case_when(
        respiratory.breath_diarrhoea_fatigue_fever_numeric == 0 ~ 0,
        covid_odds_ratio_50_numeric == 0 ~ 0,
        covid_odds_ratio_50_numeric == 1 ~ 1
      )
  )
